---
title: 'Extreme Value Analysis: precipitation extremes'
author: "MMB"
date: "19/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Extremal types theorem

Let $X_{1}, \cdots. X_{n} \overset{i.i.d.}{\sim} F$ and $M_{n} = max\left\{X_{1}, \cdots, X_{n}\right\}$. Then if there exist sequences of constants $a_{n}$ >0 and $b_{n}∈R$ such that, as $n→∞$.

$$\mathbb{P}\left( \frac{M_{n}-b_{n}}{a_{n}} \leq x \right) \overset{d}{\rightarrow} G(x)$$
for some non-degenerate distribution $G$, then $G$ takes the following form:
$$G(x)=\exp\left \{ -\left( 1+\xi x \right)_{+}^{-\frac{1}{\xi}} \right\}
$$


where $x_{+}=\max(x,0)$ and $ξ$ is a shape parameter determining the rate of tail decay, with

$ξ$ > 0 giving the heavy-tailed (`Frechet`) case

$ξ$ = 0 giving the light-tailed (`Gumbel`) case

$ξ$ < 0 giving the bounded-tailed (`Weibull`) case

## Block Maxima Approach

### Determine the block size and compute maxima for blocks

`Data`: $X_{1}, X_{2}, \cdots, X_{nm}$ (e.g. daily precipitation).

`Block maxima`: $M_{i} = max\{X_{(i-1)n+1}, \cdots, X_{in}\}, \, i = 1, \cdots, m$

For example, if we take $n$=365 then we get `annual maxima` of daily precipitation of $m$ years.

### Fit a GEV to the maxima and assess fit

We assume  $M_{1}, \cdots, M_{m}$ follows a GEV distribution $GEV(\mu_{n}, \sigma_{n}, \xi)$ and we maximize the log-likelihood 

$$-\sum_{i=1}^{m} y_{i}^{-\frac{1}{\xi}}-m log(\sigma) - (\frac{1}{\xi}+1)\sum_{i=1}^{m}log(y_{i})$$
where $y_{i} = \left[1-\xi \frac{m_{i}-\mu}{\sigma}\right]_{+}$

### Perform inference for return levels, probabilities, etc

If we fit GEV to `annual maxima`, then the `r-year return level` is the value whose probability of exceedance is $p = \frac{1}{r}$ in any given year. 

$$m_{p} = \left\{ \begin{array}{ll}
\mu-\frac{\sigma}{\xi}\Big[1-\{-\log(1-p)^{-\xi}\}] \qquad \xi \neq 0;\\
\mu-\sigma log(-log(1-p)) \qquad \xi = 0. \end{array} \right.$$
## Fort Collins daily precipitation
We analyze the daily precipitation amounts (inches) from a single rain gauge in Fort Collins, Colorado. (Source Colorado Climate Center, Colorado State University <http://ulysses.atmos.colostate.edu>).

## Exploratory Data Analysis

```{r setup_packages, include=TRUE}
library(extRemes)
data(FCwx)
head(FCwx)
```

## Exploratory Data Analysis
```{r exploratory, include=TRUE}
t <- c(rep(1:365, 4)/365, rep(c((1:366)/366,
          rep((1:365)/365, 3)), 24))
plot(t + FCwx$Year, FCwx$Prec, type = "l",
     xlab = "Year", ylab = "Daily Precip (0.01 in)")
days_in_yr <- c(rep(365, 4), rep(c(366,
          rep(365, 3)), 24))
#Extract annual max precip
annmax <- aggregate(FCwx$Prec, by = list(FCwx$Year), max)
#Extract the timing of annual max
annmax_t <- aggregate(FCwx$Prec, by = list(FCwx$Year), function(z){
  which(z == max(z))[1]
})
max_t <- annmax_t[ ,2]/days_in_yr + 1900:1999
points(max_t, annmax[ ,2], col = "blue", pch = 16,
       cex = log(annmax[ ,2]/100 + 1))
```

## Fit GEV to Block Maxima  
```{r BM, include=TRUE}
gevfit1 <- fevd(annmax[ ,2])
gevfit1
#Diagnostic plots
plot(gevfit1)
```

Suppose we are interested in estimating 50-year return level
```{r evd, include=TRUE}
library(evd)
CI_delta <- ci(gevfit1, return.period = 50, verbose = T) 
CI_delta
CI_prof <- ci(gevfit1, method="proflik", xrange = c(300, 600),
   return.period = 50, verbose = TRUE)
CI_prof 
hist(annmax[ ,2], 15, col = "lightblue",
     xlim = c(0, 600), prob = T, ylim = c(0, 0.01),
     xlab = "annual max (0.01 in)",
     main = "95% CI for 50-yr RL")
xg <- seq(0, 600, len = 1000)
mle <- gevfit1$results$par
lines(xg, dgev(xg, loc = mle[1], 
               scale = mle[2], shape = mle[3]))
for (i in 1:3) abline(v = CI_delta[i], lty = 2, col = "blue")
for (i in 1:3) abline(v = CI_prof[i], lty = 3, col = "red")
legend("topleft", legend = c("Delta", "Prof"),
       col = c("blue", "red"), lty = c(2, 3))
```
## Temporal nonstationarity (in annual max)?
```{r nonstationarity_evd, include=TRUE}
gevfit2 <- fevd(annmax[ ,2], data = annmax,
 location.fun = ~annmax[ ,1])
lr.test(gevfit1, gevfit2)
```
# Peak Over Threshold Approach

If $M_i$ is approximately $GEV(\mu, \sigma, \xi)$, then, for a large $u$
$$\mathbb{P}\left(X_{i}-u > y| X_{i} >u\right) \approx \left(1+\xi \frac{y}{\tilde{\sigma}}\right)_{+}^{-\frac{1}{\xi}}$$
  When $\tilde{\sigma} = \sigma + \xi(u - \mu)$ and $y>0$. 
  It is the survival function for the `Generaliaze Pareto distribution` (GPD)
  ```{r pot, include=TRUE}
plot(1:36524, FCwx$Prec, type = "l",
     xlab = "Year", ylab = "Daily Precip (0.01 in)",
     xaxt = "n")
yr <- c(rep(365, 3), rep(c(366, 365, 365, 365), 24))
axis(1, at = cumsum(yr)[seq(1, 100, len = 10)], 
     label = seq(1, 100, len = 10) + 1900)
#Threshold exceedances
thres <- 40
ex <- FCwx$Prec[FCwx$Prec >= thres]
#Extract the timing of POT
ex_t <- which(FCwx$Prec >= 40)
abline(h = 40, col = "blue", lty = 2)
points(ex_t, ex, col = "blue", pch = 16,
       cex = log(ex/100 + 1))
       ```
## Fit GPD to Threshold Exceedances
### How to choose the threshold?


```{r pot_plot, include=TRUE}
mrlplot(FCwx$Prec)
threshrange.plot(x = FCwx$Prec, r = c(20, 60))
gpdfit1 <- fevd(FCwx$Prec, threshold = 40, type = "GP")
gpdfit1
plot(gpdfit1)
```

Again we are interested in estimating 50-year return level
```{r RP, include=TRUE}
CI_delta <- ci(gpdfit1, return.period = 50, verbose = T)
CI_delta
CI_prof <- ci(gpdfit1, method="proflik", xrange = c(300, 600),
   return.period = 50, verbose = TRUE)
CI_prof
hist(ex, 50, col = "lightblue",
     xlim = c(40, 600), prob = T, ylim = c(0, 0.033),
     xlab = "Threshold excess (0.01 in)",
     main = "95% CI for 50-yr RL")
xg <- seq(40, 600, len = 1000)
mle <- gpdfit1$results$par
lines(xg, dgpd(xg, loc = 40, 
               scale = mle[1], shape = mle[2]))
for (i in 1:3) abline(v = CI_delta[i], lty = 2, col = "blue")
for (i in 1:3) abline(v = CI_prof[i], lty = 3, col = "red")
legend("topleft", legend = c("Delta", "Prof"),
       col = c("blue", "red"), lty = c(2, 3))
```
### Decluster the exceedances
```{r Decluster, include=TRUE}
plot(366:730, FCwx$Prec[366:730], type = "l",
     xlab = "Time", ylab = "Daily Precip (0.01 in)",
     main = "1900")
thres <- 40
ex <- FCwx$Prec[FCwx$Prec >= thres]
#Extract the timing of POT
ex_t <- which(FCwx$Prec >= 40)
abline(h = 40, col = "blue", lty = 2)
points(ex_t, ex, col = "blue", pch = 16,
       cex = log(ex/100 + 1))
####
declus <- decluster(FCwx$Prec, threshold = 40)
plot(declus)
gpdfit2 <- fevd(declus, threshold = 40, type = "GP")
CI_delta <- ci(gpdfit2, return.period = 50, verbose = T)
CI_delta
CI_prof <- ci(gpdfit2, method="proflik", xrange = c(300, 600),
   return.period = 50, verbose = TRUE)
CI_prof
hist(ex, 50, col = "lightblue",
     xlim = c(40, 600), prob = T, ylim = c(0, 0.033),
     xlab = "Threshold excess (0.01 in)",
     main = "95% CI for 50-yr RL")
xg <- seq(40, 600, len = 1000)
mle <- gpdfit2$results$par
lines(xg, dgpd(xg, loc = 40, 
               scale = mle[1], shape = mle[2]))
for (i in 1:3) abline(v = CI_delta[i], lty = 2, col = "blue")
for (i in 1:3) abline(v = CI_prof[i], lty = 3, col = "red")
legend("topleft", legend = c("Delta", "Prof"),
       col = c("blue", "red"), lty = c(2, 3))
```